---
title: "The Leslie Matrix - revisited"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

A population of fish we're interested in has four life stages: eggs (E), fry (F), juvenile (J), breeding adult (A). You are told that annually:

-   Each breeding adult will produce, on average, 600 eggs
-   5% of eggs will survive to become fry
-   20% of fry will survive to become juveniles
-   16% of juveniles will survive to become adults
-   Adult survival rate year to year is 90%

Recall the Leslie matrix from EDS 212. Matrix multiplication (using the dot product `%*%` of the matrix with the populations at each life stage) yields the population structure at the following year.

### Create the Leslie matrix:

```{r}
# Make the Leslie Matrix:
# This is the system of linear equations turned into a matrix
fish_leslie <- matrix(c(0, 0, 0, 600, 0.05, 0, 0, 0, 0, 0.20, 0, 0, 0, 0, 0.16, 0.9), nrow = 4, ncol = 4, byrow = TRUE)

# Check it out: 
fish_leslie
```

The initial population structure is given by:

```{r}
# Initial population structure (Roe, Fry, Juvenile, Adult): 
fish_year0 <- c(5000, 8000, 600, 400)
```

Add line-by-line comments wheres prompted in the code below, explaining what each is doing. Make sure that you are talking with your neighbors about what is really being created at each step.

```{r}
# Model projections

# ------ WHAT HAPPENS HERE? ------ #
# ------ comment line-by-line ---- #

# The line below... creates a vector sequence spanning from 0 to 8 years, by increments of 1 year
time_yr <- seq(from = 0, to = 8, by = 1)

# The line below... creates an empty matrix to be populated. It is 9 rows (year 0 to 8), and 4 columns (4 life stages)
proj_year <- matrix(ncol = length(fish_year0), nrow = length(time_yr))

# The line below...fills the first ROW of the empty matrix (proj_year), using the initial population data from fish_year0
proj_year[1, ] <- fish_year0

# The for loop below...starting at i = 2, up to the length in time_year (2:9), populate the matrix using a dot product. Dot product is between the formula matrix (fish_leslie) * the previous row of the proj_year matrix (hence i-1). This matrix is being filled in by row. 
for (i in 2:length(time_yr)) {
  proj_year[i,] <- fish_leslie %*% proj_year[i-1,]
}


# The rest is wrangling & visualization (run to check it out):
  # reassign column names, defaults as ([,1], [,2], etc.)
colnames(proj_year) <- c("eggs", "fry", "juvenile", "adult")

# create data frame with the years 0-9, along with the populations of each life stage
proj_df <- data.frame(time_yr, proj_year) %>% 
# melting from wide format into tall format
  #excluding time_yr (hence -time_year), take all of the column names & combine into lifestage
  pivot_longer(cols = -time_yr, names_to = "lifestage", values_to = "stage_population")

# create plot
  # time on x-axis, population stage on y-axis
ggplot(data = proj_df, aes(x = time_yr, y = stage_population)) +
# make each life stage a unique color
  geom_line(aes(color = lifestage)) +
# scale values by log 10
  scale_y_log10()

```

Test different combinations of survival / reproduction and starting population structures to see what happens!

```{r}
# Example - pivot longer
demo <- data.frame(year = c(1995, 2000, 2005))


#plotting without using pivot_longer is not ideal, you need to do two geom lines
ggplot(data = demo) +
  geom_line(aes(x = year, y = cats)) +
  geom_line(aes(x = year, y = dogs))


# plotting 
# melting cats and dogs together, so they can all be on the y axis together (without two geom lines)
# the names_to renames the new y-value, which combines cats and dogs
demo_longer <- demo %>% 
  pivot_longer(cols = c(cats,dogs), names_to = "species", values_to = "count")
```


